---
output: github_document
editor_options:
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# colvR

<!-- badges: start -->
```{r badges, echo=FALSE, results='asis'}
user <- "BBricout"   # mets ici exactement ton identifiant GitHub
repo <- "colvR"
cat(sprintf(
  "[![R-CMD-check](https://github.com/%s/%s/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/%s/%s/actions/workflows/R-CMD-check.yaml)",
  user, repo, user, repo
))
```

[![R-CMD-check](https://github.com/BBricout/colvR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BBricout/colvR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

colvR provides modeling tools for count data — especially overdispersed data with many zeros — and includes missing-data imputation. It also exposes fast Rcpp/Armadillo utilities for matrix–vector conversions and low-rank ZIP/PLN optimization.

## Installation

From GitHub:

``` r
# install.packages("devtools")
devtools::install_github("BBricout/colvR")

library(colvR)
```

When on CRAN, you’ll be able to do: install.packages("colvR")

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(colvR)

n <- 1500 ; p <- 30 ; q <- 4 ; d <- 4
dim <- list(n = n, p = p, d = d, q = q)
X <- cbind(rep(1, n*p), rnorm(n*p), rep(rnorm(n), p))
trend <- rep(1:p, each = n)
X <- cbind(X, scale(trend))
B <- c(2, 0.7, 0.5, -0.6) ; D <- c(-0.5, 0.5, 0.4, -0.4) ; C <- matrix(rnorm(p*q)/2, nrow = p, ncol = q)
theta <- list(B = B, D = D, C = C)

sim <- Simul(X, theta, dim) ; Y <- sim$Y

fit <- Miss.ZIPLNPCA(Y, X, q)
```


