---
title: "Getting started with colvR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with colvR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
# Si torch était chargé par un profil/IDE, on le décharge proprement.
if ("torch" %in% loadedNamespaces()) try(unloadNamespace("torch"), silent = TRUE)
```


The **`colvR`** package provides tools for the analysis of multivariate count data that exhibit both 
**overdispersion** and **zero-inflation**, two common features in ecological and environmental datasets.  
In particular, the package makes it possible to **impute missing data**, to estimate **Poisson log-normal models** 
enriched with a **latent PCA component**, and to optimize model parameters through efficient numerical routines.

```{r}
library(colvR)
```

## General model

We consider a dataset of counts collected at \(n\) sites across \(p\) years.  
Let \(Y_{ij}\) denote the number of individuals recorded at site \(i \in \{1,\dots,n\}\) and year \(j \in \{1,\dots,p\}\).  
Each observation is associated with a covariate vector \(x_{ij} \in \mathbb{R}^d\).  

The full covariate matrix \(X \in \mathbb{R}^{np \times d}\) typically combines three types of information:

- **Site-level covariates** (\(R \in \mathbb{R}^{n \times d_1}\)): descriptors constant across years (e.g. habitat type, geographic attributes).  
- **Year-level covariates** (\(C \in \mathbb{R}^{p \times d_2}\)): descriptors varying across years but shared across sites (e.g. climate indices, survey effort).  
- **Site–year covariates** (\(E \in \mathbb{R}^{np \times d_3}\)): descriptors specific to each site–year combination (e.g. water level, vegetation cover, observer identity).  

These blocks are concatenated into:
\[
X = [R \otimes \mathbf{1}_p \;\;\; \mathbf{1}_n \otimes C \;\;\; E], 
\qquad d = d_1 + d_2 + d_3,
\]
so that each observation \((i,j)\) is described by a covariate vector \(x_{ij}\).  

---

The **Zero-Inflated Poisson Log-Normal with Probabilistic PCA (ZI-PLN-PCA)** model generates abundances in two steps:

1. **Presence/absence process**
\[
U_{ij} \sim \text{Bernoulli}\!\left(\pi_{ij}\right), 
\qquad \pi_{ij} = \frac{1}{1+\exp(-x_{ij}^\top \gamma)},
\]
where \(U_{ij}=0\) deterministically yields \(Y_{ij}=0\).  
The parameter vector \(\gamma\) encodes how covariates affect the probability of presence.  

2. **Abundance process conditional on presence**
\[
Y_{ij} \;\sim\; \text{Poisson}\!\left(\exp(x_{ij}^\top \beta + Z_{ij})\right),
\]
where \(\beta\) controls the covariate effects on abundance, and \(Z_{ij}\) is a latent Gaussian effect.  

Each site \(i\) is associated with a \(q\)-dimensional latent vector
\[
W_i \sim \mathcal{N}(0,I_q), 
\qquad Z_i = C W_i, \quad C \in \mathbb{R}^{p \times q},
\]
so that the covariance structure of the latent effects is
\[
\Sigma = C C^\top .
\]

This **low-rank formulation** parallels probabilistic PCA and parsimoniously captures correlations between years within the same site.

## Example

### Data

To illustrate the use of `colvR`, we rely on the dataset **`fuligule_milouin`** provided with the package.  
This dataset contains winter count data of the **Common Pochard (*Aythya ferina*)**, a Palearctic diving duck species.  
The counts were collected across multiple wetland sites in North Africa over several years, and the dataset is  
characterized by a high proportion of zeros and a substantial amount of missing values — making it a suitable  
case study for the zero-inflated Poisson log-normal framework implemented in `colvR`. 



```{r example}

data("fuligule_milouin")
Y <- fuligule_milouin$Y ; X <- fuligule_milouin$X
n <- nrow(Y) ; p <- ncol(Y) ; d <- ncol(X)

cat(mean(is.na(Y)), "is the proportion of missing entries in Y.\n")
cat(mean(Y == 0, na.rm = TRUE), "is the proportion of 0's in Y.\n")


```

### Choosing the size of the latent space

An important step in model specification is the choice of the dimension \(q\) of the latent space.  
The function `select_model_bic()` can be used to compare models fitted with different values of \(q\),  
based on the Bayesian Information Criterion (BIC). The selected value corresponds to the latent space  
dimension that maximizes the BIC score.  

We rely on precomputed BIC values (saved in the package) to keep the vignette fast.
If you want to recompute them locally, set `Sys.setenv(NOT_CRAN = "true")` before knitting.

```{r}
q_grid <- 1:p

use_precomputed <- !identical(Sys.getenv("NOT_CRAN"), "true")

if (use_precomputed) {
  bics <- readRDS(system.file("extdata", "bics_fuligule.rds", package = "colvR"))
} else {
  bics <- select_model_bic(Y, X, Miss.ZIPLNPCA, q_grid)
  # (Optionnel) sauvegarder à nouveau si vous voulez mettre à jour le package :
  # saveRDS(bics, file = "inst/extdata/bics_fuligule.rds", compress = "xz")
}

# Selected q (robuste aux noms)
q <- if (!is.null(names(bics))) as.integer(names(bics)[which.max(bics)]) else q_grid[which.max(bics)]
q
```

```{r bics_plot, fig.cap="Model selection via BIC", fig.alt="Line plot of BIC versus the latent dimension q for the fuligule_milouin data; the selected q maximizes BIC."}
plot(q_grid, as.numeric(bics), type = "b",
     xlab = "Latent dimension q", ylab = "BIC")
# (optionnel) surligner q*
points(q, bics[which.max(bics)], pch = 19, cex = 1.2)
abline(v = q, lty = 2)
```


### Fit

Once the latent dimension \(q\) has been chosen, we can fit the model using  
the function `Miss.ZIPLNPCA()`. This function estimates the model parameters  
while simultaneously imputing the missing entries in the count matrix `Y`.  

```{r}
fit <- Miss.ZIPLNPCA(Y, X, q)
```

The returned object is a list containing:  

- the estimated **parameters** of the model (`beta`, `gamma`, `C`, …),  
- the **variational parameters** used for inference,  
- and the **imputed count matrix**, where missing values in `Y` are replaced by their predictions.  

This fitted object can then be explored to inspect estimated parameters, check convergence,  
and evaluate the quality of imputations.

### Confidence interval


Beyond point estimates, `colvR` also provides tools to quantify uncertainty in the parameters.  
The function `V_theta()` computes an estimate of the covariance matrix of the parameters,  
based on the sandwich estimator. Using this, we can derive approximate confidence intervals  
for the regression coefficients of both the presence (`γ`) and the abundance (`β`) models.  


```{r}
cov <- V_theta(Y, X, fit)
theta <- c(fit$mStep$gamma, fit$mStep$beta)
var <- diag(cov)[1:(2*d)]
Intervals <- IC(theta, var)
```


We now display the point estimates and 95% confidence intervals for the regression
coefficients of the presence model (\(\gamma\)) and the abundance model (\(\beta\)).

```{r ci_forest_plot, fig.cap="Regression coefficients with 95% confidence intervals", fig.alt="Horizontal dot-and-whisker plot showing gamma (presence) and beta (abundance) coefficients with their 95% confidence intervals; a vertical line marks zero."}
# Build a compact table of estimates and 95% CIs (no dependency on IC())
se <- sqrt(var)
par_names <- c(paste0("gamma_", seq_len(d)), paste0("beta_", seq_len(d)))
ci_tab <- data.frame(
  param = par_names,
  estimate = theta,
  lo = theta - 1.96 * se,
  hi = theta + 1.96 * se,
  group = rep(c("gamma (presence)", "beta (abundance)"), each = d),
  stringsAsFactors = FALSE
)

# Order for plotting: gamma first, then beta, top-to-bottom
ci_tab$param_pretty <- factor(ci_tab$param, levels = rev(ci_tab$param))

op <- par(mar = c(5, 10, 2, 2))
plot(
  x = ci_tab$estimate, y = as.numeric(ci_tab$param_pretty),
  xlab = "Coefficient (estimate with 95% CI)", ylab = "",
  yaxt = "n", pch = 19, xaxs = "i", col = ifelse(ci_tab$group == "gamma (presence)", "black", "gray30")
)
segments(ci_tab$lo, as.numeric(ci_tab$param_pretty),
         ci_tab$hi, as.numeric(ci_tab$param_pretty),
         lwd = 2, col = ifelse(ci_tab$group == "gamma (presence)", "black", "gray30"))
abline(v = 0, lty = 2)
axis(2, at = seq_along(levels(ci_tab$param_pretty)), labels = levels(ci_tab$param_pretty), las = 1)
legend("bottomright", inset = 0.01, bty = "n",
       legend = c("gamma (presence)", "beta (abundance)"),
       pch = 19, col = c("black", "gray30"))
par(op)

```


















